
---
title: 'Project 1: Wrangling, Exploration, Visualization'
author: "SDS322E"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```

## Data Wrangling, Exploration, Visualization

### Alvin Kalathungal | aak2874

#### Introduction 
```{r}
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
#install.packages("maptools")
library(maptools)
#install.packages("mapdata")
library(mapdata)
#install.packages("ggthemes")
#install.packages("usmap")
library("usmap")
library(ggthemes)
#install.packages("tibble")
library(tibble)
#install.packages("viridis")
library(viridis)
#install.packages("readr")
library(readr)
```

Paragraph or two introducing your datasets and variables, why they are interesting to you, etc.

```{R}

USVaccineData <- read_csv("USVaccineData")
#head(USVaccineData)
USCovidData <- read_csv("USCounties.csv")
#head(USCounties)

```

#### Tidying: Reshaping

If your datasets are tidy already, demonstrate that you can reshape data with pivot wider/longer here (e.g., untidy and then retidy). Alternatively, it may be easier to wait until the wrangling section so you can reshape your summary statistics. Note here if you are going to do this.

```{R}
# your tidying code (if applicable; can also wait until wrangling section)
```

    
#### Joining/Merging

```{R}
#nrow(USCovidData) #3249
#nrow(USVaccineData) #3282
#distinct(USCovidData)
#distinct(USVaccineData)
#left_join(USCovidData,USVaccineData, by="fips") %>% distinct(fips,.keep_all = TRUE) -> projectData  #3217
#anti_join(USVaccineData,USCovidData, by="fips") #10 that is in Vaccine but not Covid Data c(cities, bourgough)
#anti_join(USCovidData,USVaccineData, by="fips")  # 2 in Northern Mariana Islands
```

Discussions of joining here. Feel encouraged to break up into more than once code chunk and discuss each in turn.

####  Wrangling

```{R}
projectData %>% group_by(state) %>% summarise_at(c("cases","deaths"),.funs =list(mean=mean,sd=sd) ) %>% pivot_longer(-1) %>% separate(name, into = c("var","stat")) %>% pivot_wider(names_from ="var", values_from = "value") %>% slice(1:25) %>% kable()
```
```{r}
projectData %>% group_by(state) %>% count(state,cases,deaths) %>% summarise(totalcases = sum(cases), totaldeaths = sum(deaths)) %>% mutate(percentSurvival = 1 - totaldeaths/totalcases) %>% arrange(desc(totalcases)) %>% slice(1:10) %>%  ggplot() + geom_point(aes(x=totalcases, y=totaldeaths,color=state)) + scale_x_continuous(labels = scales::comma)
```


```{r}
projectData %>% group_by(Metro_status,state) %>% summarise_at("cases",.funs = list(mean=mean,sd=sd)) %>% arrange(state) %>% filter(state == "California" | state == "Texas"| state == "New York" | state == "Pennsylvania"| state == "Illinois"| state == "Ohio"| state == "Georgia" | state == "North Carolina" | state == "Michigan" | state == "Florida") %>% ggplot(aes(x=state,y=mean, fill=Metro_status), ) + geom_bar(stat = "summary",position = "dodge") +scale_y_continuous(breaks = seq(0,600000,10000),labels = scales::comma) + theme(axis.text.x=element_text(angle=60, hjust=1,vjust = 1)) 
```
```{r}
projectData %>% group_by(state) %>% count(state,cases) %>% summarise(totalcases = sum(cases))  -> CovidtotalStateCases

plot_usmap(data = CovidtotalStateCases, values = "totalcases", color="red" ) +  scale_fill_continuous(low = "white", high = "orangered",name = "Total Cases by State", label = scales::comma) + theme(legend.position = "right") + ggtitle("Total COVID-19 cases in US") + theme_dark() +  theme(axis.title.x=element_blank(),axis.text.x=element_blank(), axis.ticks.x=element_blank()) + theme(axis.title.y=element_blank(),axis.text.y=element_blank(), axis.ticks.y=element_blank())
```
```{r}
plot_usmap(regions = "counties", data = projectData, values = "cases", include = c("TX")) + scale_fill_continuous(low = "white", high= "red",name = "Total Cases by County", label = scales::comma) + theme(legend.position = "left") + ggtitle("Cases by County in Texas")
```


Your discussion of wrangling section here. Feel encouraged to break up into more than once code chunk and discuss each in turn.


#### Visualizing

```{R}
# your plot 1
```

Your discussion of plot 1

```{R}
# your plot 2
```

Your discussion of plot 2

```{R}
# your plot 3
```

Your discussion of plot 3

#### Concluding Remarks

If any!




